---
title: "Project 1 MT5767"
author: "PB, KA, HS"
date: "29/01/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(plotly)

```

## Question 1

We are given a species with four age classes. The dynamics of the species arise from  the sequence of sub-processes: survival, ageing, birth. To simulate population dynamics we have created a state space model, consisting of the state model and the observation model. 
The state model is a BAS model and takes into consideration each sub-processes effect on the age categories. It is a first order markov process, thereby is only dependent on the last population estimates. The birthing and survival processes are stochastic and depend on distributions, adding variability to the model, whereas ageing is deterministic. These state model estimates are considered as our true population sizes. 
We then create an observation model, which depends on the true population at the current time step. It considers what we would likely observe if we were to survey the population.


$\large{State~Model}$

$\large{Survival}$

A proportion of each state survives, denoted below as $u$, with each age category having a specific survival probability; it is noted that all individuals in age category 4 die. Below, $n_{i,t-1}$ denotes the population size at the previous time with $i=1,..,4$ representing the respective age category. Since only a proportion of the previous population survives with a certain probability, a binomial model was selected.

$$ u_{1(s),1,t} \thicksim binomial(n_{1, t-1}, 0.45) \\
u_{1(s),2,t} \thicksim binomial(n_{2,t-1}, 0.7) \\
u_{1(s),3,t} \thicksim binomial(n_{3,t-1}, 0.7) \\
u_{1(s),4,t} = 0 $$

$\large{Ageing}$

The population is structured such that no individual can stay in their age category between years and must age up, with the individuals dying after age 4. Hence, no individuals are in the first class after ageing due to it being the youngest category.

$$ u_{2(r),1,t} = 0 \\
u_{2(r),2,t} = u_{1(s),1,t} \\
u_{2(r),3,t} = u_{1(s),2,t} \\
u_{2(r),4,t} = u_{1(s),3,t} $$

$\large{Birth}$

In this model, only age categories 2 and 3 reproduce with their respective reproduction rates. These births enter into the population in age category 1; the number of individuals in age categories 2 and above are not effected in this sub-process. As we arE dealing with reproduction rates and not probabilities, with the reproduction rate for age 3 individuals indicating more than one offspring per some individuals, a poisson distribution was chosen for each birthing process.

$$ u_{3(b),1,t} = u_{2(r),1,t} + b_{2,t} + b_{3,t} \\
u_{3(b),2,t} = u_{2(r),2,t} \\
u_{3(b),3,t} = u_{2(r),3,t} \\
u_{3(b),4,t} = u_{2(r),4,t} \\
where~\small{b_{2,t} \thicksim poisson(0.9*u_{2(r),2,t}) \\
b_{3,t} \thicksim poisson(1.9*u_{2(r),3,t})} $$


Hence, from the above sub-processes, our state model produces species abundance estimates, $n_{i, t}$ for each age category $i$ at time $t$:

$$ n_{i,t} = u_{3(b),i,t} ~~~~~~ \small{for~ i=1,...,4} $$

$\large{Observation~Model}$

Finally, we want to generate our observation model; this is the number of individuals we would see in a survey taken at time $t$. It is stated there is no double counting, hence every observation estimate should be smaller than the state process model abundance estimate $n_{i,t}$. By this methodology, a binomial distribution was chosen, with each age category having the same probability of detection:

$$ y_{i,t} \thicksim binomial(n_{i,t}, 0.5) ~~~~~~
\small{for~i=1,...,4}$$

Coding these models, a function was created for a four age species, which took inputs: initial population; survival rates; birth rates; probability of detection; number of years to simulate. The output is a data frame of simulated observation and state model estimates.

```{r}

# inputs
  # n0 - vector of initial population abundances by age
  # phi - vector of survival probabilities for each age
  # rho - vector of birth rates for each age
  # detect_p - probability of detecting an individual
  # nyears - number of years to simulate 
# the function assumes that no age 4 individuals survive and no individuals age into age 1
bas_ssm <- function(n0, phi, rho, detect_p, nyears) {
  
  # define matrix for state model variables
  # column for each age category; row per year
  state_mat <- matrix(0, nrow = nyears, ncol = length(n0))
  # add initial data
  state_mat[1, ] <- n0
  
  # define matrix for observation model variables
  obs_mat <- matrix(0, nrow = nyears, ncol = length(n0))
  # add initial observation
  obs_mat[1, ] <- rbinom(4, state_mat[1, ], detect_p)
  
  # loop through our state space BAS model and observation model to simulate estimates for number of years specified 
  for (i in 2:nyears) {
    # survive 
    u1_1 <- rbinom(1, state_mat[i - 1, 1], phi[1])
    u1_2 <- rbinom(1, state_mat[i - 1, 2], phi[2])
    u1_3 <- rbinom(1, state_mat[i - 1, 3], phi[3])
    # u1_4 = 0
    
    # age
    # u2_1 = 0
    u2_2 <- u1_1
    u2_3 <- u1_2
    u2_4 <- u1_3
    
    # birth
    u3_1 <- rpois(1, rho[2] * u2_2) +
      rpois(1, rho[3] * u2_3)
    
    # overall
    state_mat[i, ] <- c(u3_1, u2_2, u2_3, u2_4)
    
    
    # observation model
    obs_mat[i, ] <- rbinom(4, state_mat[i, ], detect_p)
  }
  return(data.frame(Year = 1:nyears,
                    Abundance = c(state_mat, obs_mat),
                    Type = rep(c("State", "Observation"), each = nyears * 4),
                    State = rep(c("Age 1", "Age 2", "Age 3", "Age 4"), each = nyears,
                                times = 2)))
}

```

We are told our initial population sizes are:
$$ n_{1,1} = 150 \\
n_{2,1} = 70 \\
n_{3,1} = 50 \\
n_{4,1} = 30$$

The variables were coded for the above function, using our probabilities and rates discussed above, as well as these initial population sizes and the function was run to simulate 25 years.

```{r}

n0 <- c(150, 70, 50, 30)
phi <- c(0.45, 0.7, 0.7, 0)
rho <- c(0, 0.9, 1.9, 0)

# set seed for replicability
set.seed(2791)

model <- bas_ssm(n0, phi, rho, 0.5, 25)

```

Looking at the results graphically, we consider the overall population dynamics, combining all age groups.

```{r, echo=F, message=F}

ov_abundance <- model %>%
  group_by(Type, Year) %>%
  summarise(`Overall Abundance` = 
              sum(Abundance)) %>%
  ggplot(aes(x = Year, y = `Overall Abundance`, col = Type)) +
  geom_line() +
  labs(col="Model") +
  ylab("Abundance") + 
  ggtitle("Figure 1.1 - Overall Abundance")

ggplotly(ov_abundance, ttoltip = c("Overall Abundance", "Year"))

```

Then, splitting the models by age group.

```{r, echo=F}

age_abundance <- ggplot(model, aes(x = Year, y = Abundance, col = Type)) +
  geom_line() +
  facet_wrap(~ State) +
  labs(col="Model") +
  ggtitle("Figure 1.2 - Abundance by Age")

ggplotly(age_abundance, ttoltip = c("Abundance", "Year"))

```

We can see from figures 1.1 and 1.2, that the age category which is the most similar to the overall abundance shape is age 1; this is expected due to this age class contributing most to the overall population. At the start, age 1 makes up half of the population!

After age 1, where the probability of survival to age 2 is below half,  each category gradually declines in size. Age 3 births' contribute the most to age 1 as its birth rate ($\rho_3 = 1.9$) is more than double that of age 2 ($\rho_2 = 0.9$). Hence, when we see a large increase from one year to the next in age 1 there is often an increase between those years in age 3 individuals - this can be seen between years 7 and 8 where there is a large increase in age 1 individuals which can also be seen in age 3, whereas in age 2 the number of individuals actually decreases between these years.

Comparing our state model to our observation model, we usually see around half of the actual population, which is to be expected with our probability of detection $p = 0.5$.

## Question 2

Question 2 considers a wildebeest example seen in lectures, where abundance estimates were produced from aerial surveys taken in the Serengeti between 1960 and 1989. The rainfall for the season and the number of animals taken by illegal harvesting were also recorded yearly. Trying to incorporate these variables into the logistic growth model used to model the population, showing their impact on the wildebeest population, we consider different forms of the logistic growth model in this question.

### Part A
We are given the model where illegal harvesting is assumed to take place *after* growth:
  
  $$ N_t = N_{t-1} + r_tN_{t-1}(1 - \frac{N_{t-1}}{K_t} ) - c_{t-1} $$
  This is because the measurement of hunting $c$ is assumed to be measured at the end of the year, i.e by the end of year $t-1$, $c_{t-1}$ million Wildebeest have been killed and hence this will be taken off the measurement of year $t$ (ie after growth). However if we want to harvest *before* growth, we would assume $c$ represents that it knows the number of Wildebeest which will be killed at time $t$. Therefore, if we want the model to reflect this case we would use $c_t$ in the model, harvesting taking place for the current year before they grow in the next year:
  
  $$ N_t = N_{t-1} + r_tN_{t-1}(1 - \frac{N_{t-1}}{K_t} ) - c_t $$

### Part B

Thinking biologically about the situation the carrying capacity and growth rate
depend upon rainfall. There is a postive association, which indicates that
increased rainfall leads to larger amounts of grass for wildebeest to eat
leading to higher carrying capactiy and growth rate. 

There are a number of factors to consider

Wildebeest usually give birth at the start of the wet season, 
this would indicate that the current rainfall value is more important
as wildebeest generally graze on short rapidly growing grass and rainfall
in the current year would have a higher impact on growth on the rapidly growing
frequently grazed grass than the previous year, leading to a higher supply for 
young wildebeest leading to a higher growth rate. 

Additionally the fact that wildebeest have an annual migration further supports
the use of rainfall on the current year for use in the model as wildebeest 
specifically migrate in response to rainfall in the current year in relation
to wet and dry seasons. 
  
Note: will email lecturer to check if I should include sources regarding
general biological information. Still working on section. 



### Part C

The carrying capacity, $K_t$ of Wildebeest at time t is dependent on a few factors in their habitat. These include: Food supply, water supply, habitat space and competition. 
Wildebeest have a diet that consists strictly of grazing. Intuitively, this would depend on the rainfall of the current year; the effects of droughts can be very rapid and cause grass to dry out on a timescale of weeks, which may hinder the habitat's ability to support the animal population that depends on it.  

A study in the Journal of Ecology [Joan Dudney et al. 2017] suggests that the lagged rainfall may have a significant influence over the abundance of plant species in annual-dominated grasslands. They note that "the effect of previous-year precipitation was greater for forbs than grasses, indicating that rainfall-induced biotic effects
strongly constrain forb growth and weakly impact grasses." This shows that the lagged rainfall has a greater effect on non-grass plant species than grasses, and therefore has a weaker implication on the abundance of the main food source of the wildebeest. 

Another study, [Gundula S. Bartzke et al.] conducted in the Serengeti national park, notes that "the increasing rainfall extremes in the Mara can be expected to create conditions conducive to outbreaks of infectious animal diseases and reduced vegetation quality for herbivores, particularly when droughts and floods persist over multiple years'. It may therefore be useful for ecologists to incorporate the effect long drought periods can have on carrying capacity. This may be beyond the scope of the data collected, but a more in depth analysis could highlight key changes in the carrying capacity this way. 

Overall, the effects of rainfall measured in the current year have a much greater effect on the carrying capacity. There is a much clearer pattern between rainfall and the growth of vegetation at smaller timescales, which have a large impact on the habitat's ability to support the species, which would indicate that a model for $K_t$ based on $r_{t}$ will be more informative.
$$ N_t = N_{t-1} + r_{t}N_{t-1}(1 - \frac{N_{t-1}}{K_t} ) - c_{t-1} \\
K_t = \exp{(\beta_0 + \beta_1R_t)}$$

## Question 3
### Part A
In this part we will be fitting two formulation of the time varying growth rate
model. One where $r_t$ (growth rate over time) is modeled as a function of rain
in the same year $R_t$ and another where $r_t$ is modeled as a function of 
rain in the previous year. As seen previously the log linear model of the form 
$$ N_t = N_{t-1} + r_{t}N_{t-1}(1 - \frac{N_{t-1}}{K_t} ) - c_{t-1} $$
is used. 

Where $N_t$ is the population at time step t, 

Where $N_t$ is the population at time step t-1, 

$K_t$ is the carrying capacity which is assumed to be fixed and 

$c_{t-1}$ is the removal from illegal hunting from the previous year. 

Specifically in this case we are concerned with the growth rate, thus we assume
constant carrying capacity. This gives us the following equations

### Equation 1 

Where the growth rate is expressed as a function of rainfall in the same year.

$$ r_t = exp(\alpha_0 +\alpha_1R_t)$$

### Equation 2

Where the growth rate is expressed as a function of rainfall in the 
previous year. 

$$ r_t = exp(\alpha_0 +\alpha_1R_{t-1})$$

### Equation 3

Where the carrying capacity is assumed to be constant. 

$$ k_t = exp(\beta_0) $$

We shall then proceed to fit the model using R. 

```{r}
library(statsecol)
data(wildebeest)

# Coding in progress using practical as a base

#defining function

rain_rK_nll <- function(pars, years, removals, Nhat, SEhat, rain, model=1, type="nll"){
  
  #parameter set up
  
  N0 <- exp(pars[1])
  N <- numeric(years)
  r <- numeric(years)
  k <- numeric(years)
  N[1] <- N0
  r[1] <- NA
  k[1] <- NA

  # Current year model 
  
  if(model == 1){
    if(length(pars)!=4){stop("par should have 4 values")}
    r[2:years] <- exp(pars[2]+pars[3]*rain[2:years])
    k[2:years] <- rep(exp(pars[4]),years-1)
  }
  
  # Previous year model 
  
  if(model == 2){
    if(length(pars)!=4){stop("par should have 4 values")}
    r[2:years] <- exp(pars[2]+pars[3]*rain[1:years-1])
    k[2:years] <- rep(exp(pars[4]),years-1)
  }
  
    #generate population dynamics:
  for(i in 2:years){
    N[i]=N[i-1] + r[i] * N[i-1] * (1-N[i-1]/k[i]) - removals[i-1]
  }
  
  negloglik <- -sum(dnorm(Nhat,N,SEhat,log=TRUE), na.rm=TRUE)
  
  #what should be returned? 
  if(type=="nll"){  return(negloglik)}
  if(type=="proj"){ return(N)}
}


#assigning parameter values based upon the data

yrs <- nrow(wildebeest)
rmv <- wildebeest$Catch
Nhat <- wildebeest$Nhat
SEhat <- wildebeest$sehat
rain <- wildebeest$rain

# fitting the two models:
fit_1 <- optim(par = c(log(0.1),log(0.25),0, 0), fn = rain_rK_nll, years = yrs, 
               removals = rmv, Nhat = Nhat, SEhat = SEhat, rain = rain, model = 1)
fit_2 <- optim(par = c(log(0.1),log(0.25),0, 0), fn = rain_rK_nll, years = yrs, 
               removals = rmv, Nhat = Nhat, SEhat = SEhat, rain = rain, model = 2)

# checking parameter values
# Because of the way we've worked our function pars[3] is our alpha 1

#model 1 alpha 1
alpha_1_1 <- fit_1$par[3]

#Model 2 alpha 1

alpha_2_1 <- fit_2$par[3]

print(alpha_1_1)

print(alpha_2_1)

#calculating AIC

aic1 <- 2*fit_1$value + 2*length(fit_1$par)
aic2 <- 2*fit_2$value + 2*length(fit_2$par)

aictab <- data.frame(
            Model = c("r(t)k(.)","r(t-1)k(.)"),
            AIC = c(aic1,aic2),
            dAIC = c(aic1,aic2)-min(c(aic1,aic2)))
aictab[order(aictab$dAIC),]

```
As can be seen from above we obtain an AIc value of -18.96683 for the current
year model and an AIC of -17.14583 from the previous year model. 
This would indicate that the current year model is preferable as it has lower AIC.

This also indicates that the slope parameter of approximately 0.63 from 
alpha_1_1 the current year model is preferable which is different from the 0.15 
slope parameter for the previous year model indicating that a steeper slope
parameter is preferable. 

```{r}

# Note this section is used for graphing and requires the previous R chunk to be
# run to work

proj_1 <- rain_rK_nll(fit_1$par, years = yrs, removals = rmv, Nhat = Nhat, 
                      SEhat = SEhat, rain = rain, model=1, type="proj")
proj_2 <- rain_rK_nll(fit_2$par, years = yrs, removals = rmv, Nhat = Nhat, 
                      SEhat = SEhat, rain = rain, model=2, type="proj")

pred_df <- data.frame(years = rep(wildebeest$year,4),
                      N = c(proj_1,proj_2),
                      Model=rep(aictab$Model,each=nrow(wildebeest)))

gpreds <- ggplot(wildebeest, aes(x=year, y=Nhat)) +
          geom_errorbar(aes(ymin=lci,ymax=uci), width=0) +
          ggtitle("Figure 3.1- Abundance over years for current year vs previous year models") +
          geom_point(size=3) +
          geom_line(data=pred_df, aes(x=years,y=N,color=Model,group=Model),size=0.8) +
          ylim(0,2.1) + ylab("Abundance (millions)") + xlab("Year") +
          theme_bw()
gpreds

```
As can be seen from the graph above both models fit well and are within error
bounds which is expected given the low AIC previously, however it can be seen
that The model represented by the blue line has a better fit as the line
is closer to the center of the error bounds for most points. It should also be 
of note that there are no large local minima or maxima on the graph, 
which is good as it indicates that parameter estimates will not lead
to false predictions. 



---


## Question 3
#### part b

As discussed in question 2 part b, we would like to find out whether it is more appropriate to model the carrying capacity of the wildebeest, $K_t$. To do this, we can make a function that returns the projection of the wildebeest population and also the negative log likelihood for each model, K_rain. 
The projection model we will use is:
$$ N_t = N_{t-1} + r_{t}N_{t-1}(1 - \frac{N_{t-1}}{K_t} ) - c_{t-1} $$
where:
$$ K_{t_1} = exp(\beta_0 + \beta_1 R_{t})$$
and 
$$ K_{t_2} = exp(\beta_0 + \beta_1 R_{t-1})$$

Keeping the growth parameter $r_t$ fixed.


``` {r}
library(statsecol)
data("wildebeest")
K_rain <- function(pars, years, removals, Nhat, Sehat, rain, model = 1, type = "proj"){

  #define variables from the inputs
  N0 <- exp(pars[1])
  N <- k <- numeric(years)
  r <- rep(exp(pars[2]), years)
  
  N[1] <- N0
  k[1] <- NA
  
  beta <- pars[c(3, 4)]
  
  for (i in 2:years){
    if(model == 1) {
      k[i] <- exp(beta[1]+beta[2]*rain[i])
    }
    if(model == 2) {
      k[i] <- exp(beta[1]+beta[2]*rain[i-1])
    }
    N[i] = N[i-1] + r[i] * N[i-1] * (1-N[i-1]/k[i]) - removals[i-1]
  }
  negloglik <- -sum(dnorm(Nhat,N,Sehat, log=TRUE), na.rm=TRUE)
  
  if(type=="nll"){ return(negloglik)}
  if(type=="proj"){ return(N)}
  
}
```
This also depends on the parameters: $N_0$, the initial population; The growth parameter, $r$; $\beta_0$ and $\beta_1$, the parameters that govern the carrying capacity.
To fit our parameters, we begin with the following estimates:
```{r, echo = TRUE}
N0 <- log(0.268)
r <- log(0.1)
beta0 <- 0.5
beta1 <- 0.5
pars <- c(N0, r, beta0, beta1)
```

We need to optimise the function in order to obtain these fitted parameters. We can use optim and the negative log likelihood function from the K_rain for this. 
```{r, eval = TRUE}

fitm1 <- optim(par = pars, 
                fn = K_rain,
                years = nrow(wildebeest), 
                removals = wildebeest$Catch,
                Nhat = wildebeest$Nhat,
                Sehat = wildebeest$sehat,
                rain = wildebeest$rain,
                model =1,
                type = "nll")

fitm2 <- optim(par = pars, 
                 fn = K_rain,
                 years = nrow(wildebeest), 
                 removals = wildebeest$Catch,
                 Nhat = wildebeest$Nhat,
                 Sehat = wildebeest$sehat,
                 rain = wildebeest$rain,
                 model =2,
                 type = "nll")

#storing the parameters in a data frame
fit <- data.frame(par1 = fitm1$par, par2 = fitm2$par)
print(fit)
```
Inspecting the output of coefficients as fitted, we can see that there is very little change in $N_0$ and $r$. The difference in the beta terms are much more significant: The second term, $\beta_0$ is diminished in the t-1 model, which means that the rainfall contributes very little to the carrying capacity in a given year t. When this value is close to 0, our model approaches the examples from practical 3, where the rainfall is not modeled at all. here, our $\beta_{0_2}$ = `r  fit$par2[3]` is similar to the $\beta$ = `r log(2.604)` found when the carrying capacity is fit as a constant. 

We can now use these outputs to create models of the projected populations in each model. 
```{r}
#creating the 
fitmod1 <- K_rain(model = 1,type = "proj", fit$par1, nrow(wildebeest), 
                  wildebeest$Catch, wildebeest$Nhat, wildebeest$sehat, 
                  wildebeest$rain)

fitmod2 <- K_rain(model = 2,type = "proj", fit$par2, nrow(wildebeest), 
                  wildebeest$Catch, wildebeest$Nhat, wildebeest$sehat, 
                  wildebeest$rain)

```


```{r,  include=TRUE, echo =FALSE}
abund <- ggplot(data = wildebeest, 
               mapping = aes(x= year, y =Nhat, ymin = Nhat - sehat, ymax = Nhat + sehat))+
         geom_pointrange()+
         theme_bw() +
         labs(x = "Year", y = "Abundance (millions)",
             title ="Figure 3.2 - Abundance of Wildebeest Populations")+
         geom_line(aes(y = fitmod1, colour = "t"))+
         geom_line(aes(y = fitmod2, colour = "t-1"))

ggplotly(abund, ttoltop = c("year", "fitmod1", "fitmod2"))
```
We can easily visualise these differences in the plot below. The low $\beta_0$ value in the t-1 model is characterised by the smooth slope sections bisected by the introduction of the removals beginning in 1977.


``` {r}

nll <- c(K_rain(model = 1,type = "nll", fit$par1, nrow(wildebeest), 
               wildebeest$Catch, wildebeest$Nhat, wildebeest$sehat, 
               wildebeest$rain),
         K_rain(model = 2,type = "nll", fit$par1, nrow(wildebeest), 
               wildebeest$Catch, wildebeest$Nhat, wildebeest$sehat, 
               wildebeest$rain))
#K is the number of parameters
K <- 4

#calculating the AIC
aic <- 2 * nll + 2 * K

AIC <- data.frame(Model = c("K ~ t", "K ~ t-1"),
                  K = K,
                  NLL = nll,
                  AIC = aic,
                  dAIC = c(aic - min(aic)))
AIC[order(aictab$dAIC),]
```
Looking at the table above, we can see that the model with the lower AIC is model 1, when the carrying capacity is modeled by the rainfall in the current year. This implies that the 